---
title: "Homework 3"
output: github_document
---

Complete the assignment below. Make sure you answer each part of the questions. Submit your responses on eCampus as both an .RMD file and a knitted .pdf file. This assignment is due by 2/4.

# Problem 1

The [Billionaires Statistics Dataset](https://www.kaggle.com/datasets/nelgiriyewithana/billionaires-statistics-dataset/data) is a dataset from Kaggle (.csv included in eCampus attachments) that contains information on the world's billionaires and the countries in which they reside. You can find more information on the dataset using the link above.

```{r}
library(readr)
Billionaires_Statistics_Dataset <- read_csv("Billionaires Statistics Dataset.csv")
str(Billionaires_Statistics_Dataset)
summary(Billionaires_Statistics_Dataset)
```

For this analysis, consider a scenario where you work for a wealth management firm that is looking to expand its operations to new countries. Your job is to understand what factors are associated with the number of billionaires in a country. Ultimately your firm will cross reference this information with industry forecasts on country GDP growth, etc. to determine which countries are most likely to have the highest growth in the number of billionaires in the next 10 years (outside the scope of this assignment). For now, you will build a model focusing on the factors associated with the number of billionaires in a country.

Before we can model this, we must summarize the data by country. We will count the number of billionaires per country and use the first observation for each country to summarize the country specific statistics. To do this we use the `tidyverse` package (note, you do not have to understand this code but if you are interested in learning more about `tidyverse` see the corresponding Module X section).


```{r}
library(tidyverse)
BillionairesByCountry <- Billionaires_Statistics_Dataset %>%                       
  select(country,
         gdp_country,
         gross_tertiary_education_enrollment,
         gross_primary_education_enrollment_country,
         life_expectancy_country,
         tax_revenue_country_country,
         total_tax_rate_country,
         population_country,
         latitude_country,
         longitude_country)%>% # select only the columns we are interested in
  group_by(country)%>% # group by country
  summarise(count_billionaires=n(),
            gdp_country = first(gdp_country),
            gross_tertiary_education_enrollment = first(gross_tertiary_education_enrollment),
            gross_primary_education_enrollment_country = first(gross_primary_education_enrollment_country),
            life_expectancy_country = first(life_expectancy_country),
            tax_revenue_country_country = first(tax_revenue_country_country),
            total_tax_rate_country = first(total_tax_rate_country),
            population_country = first(population_country),
            latitude_country = first(latitude_country),
            longitude_country = first(longitude_country))%>% # summarize by country
  mutate(gdp_country=as.numeric(gsub("\\$|\\,", "", gdp_country))) # fix a problem with GDP being a string
```

BLah Blah 
Now you have the summarized dataset `BillionairesByCountry`. Use this dataset to:

(1) Build a Poisson regression model to predict the number of billionaires per country (`count_billionaires`) using the other variables in the dataset (except for `country`). Note: It is industry standard to use a `log` transform when using large financial metrics such as `gdp_country` and `tax_revenue_country_country` as predictors so make sure you do that. Note: you will have to deal with "NAs", do so using `na.omit`.

(2) Use `step` for model feature selection.

(3) Check for overdispersion/ underdispersion and account for it in your final model if necessary.

(3) Use the `effects` library to create effects plots for the final model (remember to adjust the `fig.height` and `fig.width` chunk options so that the plots look nice).

(4) Write a summary of your methodology for your direct supervisor. You direct supervisor has a similar statistical background as you, but does not use it on a daily basis so you will need to briefly refresh them on the statistical concepts you used and explain your methodology in detail. You know from previous experience that your supervisor is will be interested in why you did not use `country` as a predictor, why you used a `log` transform for large financial metrics, why you chose to omit "NAs", and how you accounted for overdispersion/ underdispersion.

(5) Write a summary of the key findings for the VP of your firm that you report under who is particularly interested in this project. This VP also has a similar statistical background to you and your supervisor, but has not used in years. Further your VP is an extremely busy person and does not like reading long reports; however they will ask an annoying number of questions if they do not understand something. Strike a balance between including how you arrived at the results and the results themselves. Are there certain details you can footnote rather include directly in the body of the report? Are their resources on the web you can hyperlink for additional information rather than recreate the wheel (i.e. [Poisson regression - Wikipedia](https://en.wikipedia.org/wiki/Poisson_regression) ).

(6) Suppose your VP responds to your report with the email below. Write a brief response to this email (do not actually write any additional code or do any additional analysis for this part). 

"Great Analysis! 

As a follow-up, I'd be interested in extending the model to counties without billionaires so that we don't miss out on emerging opportunities. A junior analyst should be able to pull the metrics you need for most of the countries not listed here. I don't think you'll be able to use Poisson regression for this since the response variable will be 0 for all of these countries. It's been a while since I've had stats, can you remind me what model can be used for this case? 

Additionally, I'd be interested in a comparison between the models with and without the additional countries. I'm not sure how to do this (although cross validation seems to be ringing a bell). I'm sure you can figure it out; let me know what you plan to try and I'll see if jogs my memory. 

Lastly, I know metrics like "GDP" aren't going to be available for all countries (i.e. Hong Kong), but I'd like to see if we can estimate a value for these countries so we don't exclude them from consideration. I'm going to reach out to some consultants to see if they have any ideas; do you have any concerns I should bring up with them? In particular, would you want to use these estimates for training your model or just for prediction?

Thanks!"

(7) OPTIONAL (no points toward the assignment, but good practice): Write a brief response to the email below from your direct supervisor. You can be more informal since you work with your direct supervisor daily.

"Hey,

I noticed you used the "first" observation for each country to summarize the country specific statistics. I get why you did this (it should be the same for each country) but I would have just used the mean for each country. Is there a way you check and make sure the first observation is the same for each country, just to be safe? Send me your code with comments and I'll take a look.

Thanks!"


## Problem 2

The dataset `happy` in the `faraway` package is about 39 students from the University of Chicago MBA cohort.  

```{r}
library(faraway)
data(happy)
str(happy)
summary(happy)
help(happy)
```

We want to explain the effects of the other information on the happiness of the students.  The variable `happy` is a numeric variable that ranges from 0 to 10, with 10 being the happiest. This is recorded as a number, so we must first convert it an ordered factor variable using the code below:

```{r}
# This code converts the happy variable from a numeric variable to an ordered factor variable
myHappy <-happy %>% 
  mutate(happy = factor(happy, ordered = TRUE))
```

Consider the following models:

```{r}
# Ordinal Regression
library(MASS)
mod1<-polr(happy~.,data=myHappy)
summary(mod1)

# Multinomial Regression
library(nnet)
mod2<-multinom(happy~.,data=myHappy)
summary(mod2)
```

For this problem:

(1) Compare the two models summaries. You do not need to interpret each coefficient, but pick a particular variable and explain how the interpretation of the coefficient differs between the two models. What is different about these models? What is the same/ similar?

(2) Can AIC and Deviance be used to compare these two models (Hint: NO!)? Why or why not? What can be used to compare these two models (you do not need to write the code you this, just explain the methodology you would use)?

(3) Notice that we changed `happy` from a numeric variable to an ordered factor variable. What is the difference between an ordered factor variable and a numeric variable? Why is it important to use an ordered factor variable for `happy` in this case? Are there other variables that can benefit from a similar conversion? If so, what are they and what do they need converted to? (You do not need to write any code for this question, just answer conceptually). 

(4) Create effects plots for each model (remember to adjust the `fig.height` and `fig.width` chunk options so that the plots look nice). What do you notice about the effects plots? What is different between these two models? What is the same? Are the effects easier or harder to interpret than the coefficients? Why or why not?

(5) Which model do you prefer for this case? Why?

(6) OPTIONAL (no points toward the assignment, but good practice): Write the code for the model comparison you described in (2).

(7) OPTIONAL (no points toward the assignment, but good practice): Convert the variables you described in (3) to the appropriate variable type. Create a new model using the converted variables and compare it to the original model.