---
title: "Assignment 5"
output: github_document
date: "2024-02-24"
---

## The Data

For assignment 5, you will be looking at the [Hotel booking demand](https://www.kaggle.com/datasets/jessemostipak/hotel-booking-demand) from Kaggle (csv attached for convince). This data set contains booking information from two hotels in Prague, "Resort Hotel" and "City Hotel." 

```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(readr)
hotel_bookings <- read_csv("hotel_bookings.csv",
                           col_types = cols(reservation_status_date = col_date(format = "%Y-%m-%d")))

summary(hotel_bookings)
str(hotel_bookings)


hotel_bookings <- read_csv("hotel_bookings.csv",
                           col_types = cols(reservation_status_date = col_date(format = "%Y-%m-%d")))

summary(hotel_bookings)
str(hotel_bookings)

library(lubridate)

# Clean data set
myHotel_bookings<-hotel_bookings%>%
  mutate(arrival_date=paste0(arrival_date_day_of_month,
                             arrival_date_month,
                             arrival_date_year)%>%
           dmy(),
         arrival_status_diff=arrival_date-reservation_status_date,
         reservation_status_date_month=month(reservation_status_date),
         reservation_status_date_year=year(reservation_status_date),
         arrival_date_month_num=month(arrival_date))%>% # clean dates
  mutate(hotel=as.factor(hotel),
         meal = as.factor(meal),
         country = as.factor(country),
         market_segment = as.factor(market_segment),
         distribution_channel = as.factor(distribution_channel),
         reserved_room_type = as.factor(reserved_room_type),
         assigned_room_type = as.factor(assigned_room_type),
         deposit_type = as.factor(deposit_type),
         agent = as.factor(agent),
         company = as.factor(company),
         reservation_status = as.factor(reservation_status),
         customer_type =as.factor(customer_type))%>% # convert character variables to factors
    mutate(reservation_date = arrival_date -ddays(lead_time),
         is_LMC = as.numeric(is_canceled==1 & arrival_status_diff<=30)) # engineer new variables


# Find LMCs by month
myHotel_LMCs<-myHotel_bookings%>%
  filter(is_LMC==1)%>% # select only LMCs
  group_by(hotel, reservation_status_date_year,reservation_status_date_month)%>%
  summarise(cancelations=sum(is_LMC))%>% # get total LMCs by month, year, and hotel
  arrange(hotel, reservation_status_date_year,reservation_status_date_month) # sort the data


myHotel_bookings_completed_Jul17 <- myHotel_bookings%>%
  filter(reservation_status_date<ymd("20170701") | arrival_date<ymd("20170701"))

myHotel_bookings_outstanding_Jul17 <- myHotel_bookings%>%
  filter(reservation_status_date>=ymd("20170701") &
           arrival_date>=ymd("20170701") &
           reservation_date<ymd("20170701"))%>%
  select(-is_canceled,
         -reservation_status,
         -reservation_status_date,
         -reservation_status_date_month,
         -reservation_status_date_year,
         -arrival_status_diff,
         -assigned_room_type)

myHotel_bookings_completed_Aug17 <- myHotel_bookings%>%
  filter(reservation_status_date<ymd("20170801") | arrival_date<ymd("20170801"))

myHotel_bookings_outstanding_Aug17 <- myHotel_bookings%>%
  filter(reservation_status_date>=ymd("20170801") &
           arrival_date>=ymd("20170801") &
           reservation_date<ymd("20170801"))%>%
  select(is_canceled,
         
         -reservation_status,
         -reservation_status_date,
         -reservation_status_date_month,
         -reservation_status_date_year,
         -arrival_status_diff,
         -assigned_room_type)
```

Note that 'completed' data sets are used for model training, and 'outstanding' are used for forecasting.

```{r}
library(forecast)
# July 2017

# Summarize the data
summary_completed_Jul17 <- myHotel_bookings_completed_Jul17%>%
  group_by(reservation_status_date_year,reservation_status_date_month)%>% # we should be forecasting on 'reservation_status_date' since that is the date the actual cancellation occurred. That is my bad.
  filter(reservation_status_date<ymd("20170701"))%>% # We made the assumption that once a guest checks in, their stay will be fine. That gives us some weird data on the edge for guests who have already checked in but are still in the hotel. We should filter out those guests.
  filter(reservation_date>=ymd("20150701"))%>% # along the same vein, the data set starts with check in dates in 7/2015 but some of these canceled way ahead of time giving us some data issues. We should filter out those guests.
  filter(hotel=="Resort Hotel")%>% # don't forget we work for Resort Hotel (if you are clever you can figure out how to use the City Hotel data to help you, but I'm not giving this one away)
  summarise(LMCs = sum(is_LMC), total_bookings = n()) # LMCs is our target. We will use 'total_bookings' as an external predictor. This isn't a great way to do this, since some of these bookings were past the time we would have hypothetically had data availiable forecasting. There is a better way, but I'm not giving it away.

# Convert target into time series
summary_completed_Jul17_TS = ts(summary_completed_Jul17$LMCs, start=c(2015,7), frequency=12)


```

Last minute cancellations (LMCs) are when a guest cancels their reservations on very short notice (for this assignment we will say less than or equal to 30 days before their check-in date). They are a logistical nightmare for the hotel industry, because staffing, stocking, etc. all depend on the number of guest staying in the hotel.

Let's do some basic data cleaning and take a look at the LMCs for each hotel:

```{r}

# Clean data set
myHotel_bookings<-hotel_bookings%>%
  mutate(arrival_date=paste0(arrival_date_day_of_month,
                             arrival_date_month,
                             arrival_date_year)%>%
           dmy(),
         arrival_status_diff=arrival_date-reservation_status_date,
         reservation_status_date_month=month(reservation_status_date),
         reservation_status_date_year=year(reservation_status_date),
         arrival_date_month_num=month(arrival_date))%>% # clean dates
  mutate(hotel=as.factor(hotel),
         meal = as.factor(meal),
         country = as.factor(country),
         market_segment = as.factor(market_segment),
         distribution_channel = as.factor(distribution_channel),
         reserved_room_type = as.factor(reserved_room_type),
         assigned_room_type = as.factor(assigned_room_type),
         deposit_type = as.factor(deposit_type),
         agent = as.factor(agent),
         company = as.factor(company),
         reservation_status = as.factor(reservation_status),
         customer_type =as.factor(customer_type))%>% # convert character variables to factors
    mutate(reservation_date = arrival_date -ddays(lead_time),
         is_LMC = as.numeric(is_canceled==1 & arrival_status_diff<=30)) # engineer new variables


# Find LMCs by month
myHotel_LMCs<-myHotel_bookings%>%
  filter(is_LMC==1)%>% # select only LMCs
  group_by(hotel, reservation_status_date_year,reservation_status_date_month)%>%
  summarise(cancelations=sum(is_LMC))%>% # get total LMCs by month, year, and hotel
  arrange(hotel, reservation_status_date_year,reservation_status_date_month) # sort the data

print(myHotel_LMCs)
```

## The Assignment

For this assignment, you will assume the role of an analyst for "Resort Hotel." In this position, at the end of each month you must submit a forecast of the expect number of LMCs for the next month.

Specifically, you must:

- Pretending the date is is June 30th, 2017 (i.e. you only have access to data up to June 30th, 2017), create a forecast for the number of LMCs at "Resort Hotel" for July 2017. 
- Pretending the date is is July 31st, 2017 (i.e. you only have access to data up to July 31st, 2017), create a forecast for the number of LMCs at "Resort Hotel" for August 2017. 

To help from accidentally building models with leakage I have pre-cleaned some data sets for you:

```{r}
myHotel_bookings_completed_Jul17 <- myHotel_bookings%>%
  filter(reservation_status_date<ymd("20170701") | arrival_date<ymd("20170701"))

myHotel_bookings_outstanding_Jul17 <- myHotel_bookings%>%
  filter(reservation_status_date>=ymd("20170701") &
           arrival_date>=ymd("20170701") &
           reservation_date<ymd("20170701"))%>%
  select(-is_canceled,
         -reservation_status,
         -reservation_status_date,
         -reservation_status_date_month,
         -reservation_status_date_year,
         -arrival_status_diff,
         -assigned_room_type)

myHotel_bookings_completed_Aug17 <- myHotel_bookings%>%
  filter(reservation_status_date<ymd("20170801") | arrival_date<ymd("20170801"))

myHotel_bookings_outstanding_Aug17 <- myHotel_bookings%>%
  filter(reservation_status_date>=ymd("20170801") &
           arrival_date>=ymd("20170801") &
           reservation_date<ymd("20170801"))%>%
  select(-is_canceled,
         
         -reservation_status,
         -reservation_status_date,
         -reservation_status_date_month,
         -reservation_status_date_year,
         -arrival_status_diff,
         -assigned_room_type)


```

You must only use:

- `myHotel_bookings_completed_Jul17` to train the July 2017 model 
- `myHotel_bookings_completed_Aug17` to train the August 2017 model
- `myHotel_bookings_outstanding_Jul17` as external predictor values for the July 2017 model
- `myHotel_bookings_outstanding_Aug17` as external predictor values for the August 2017 model.

## Additional Rules

In addition to only using the time-appropriate data as specified in the previous section you must also follow these rules:

- You must use the same methodology for the July 2017 and August 2017 models.
- You must only use `gam`, `prophet`, and `auto.arima` (with or without external regressors) models.
- You must try at least two modeling methodologies and submit the code for them (i.e. a `gam` and a `prophet` model is acceptable but two `auto,arima` models is not).
- You must explain your process including why you made any decisions you made, what your code is doing, etc.
- Your decisions/ models/ etc. must be defensible without reference to data you would not have access to on the hypothetical date of your analysis.
- You may use "City Hotel" data to help you construct your model, but be careful not to accidentally include it in your final forecast.


## Evaluation

We know the true value for July 2017 is `193` and the correct value for August 2017 is `135`. Given this information, I have created the following function to calculate RMSE for your models.

 
```{r}
modelTester <- function(for1707,for1708){
  return(sqrt((((for1707-193)^2+(for1708-135)^2))/2))
}
```

Recall, lower RMSE is desirable or "better."

# Assignment 5

## EDA
```{r}
summary(myHotel_bookings_completed_Aug17)
summary(myHotel_bookings_completed_Jul17)
summary(myHotel_LMCs)

ggplot(myHotel_LMCs, aes(x = hotel, y = cancelations, fill = hotel)) +
  geom_bar(stat = "identity") +
  labs(x = "Hotel", y = "Number of LMCs", title = "Comparison of LMCs by Hotel") +
  theme_minimal()
```

```{r}
myHotel_LMCs$month <- factor(myHotel_LMCs$reservation_status_date_month, levels = 1:12, labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))
ggplot(myHotel_LMCs, aes(x = month, y = cancelations, fill = hotel)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Month", y = "Number of LMCs", title = "Monthly Seasonality in LMCs") +
  theme_minimal()
```


**Breakdown of how you will be graded for this assignment:**
## (1) Creating a reasonable forecast for July 2017, following all the rules above and using proper methodology: 10%

```{r}
library(forecast)

summary_completed_Jul17 <- myHotel_bookings_completed_Jul17 %>%
  filter(reservation_status_date < ymd("20170701"),
         reservation_date >= ymd("20150701"), 
         hotel == "Resort Hotel") %>%
  group_by(reservation_status_date_year, reservation_status_date_month) %>%
  summarise(LMCs = sum(is_LMC), total_bookings = n()) %>%
  ungroup() 

summary_for_prophetJul17 <- summary_completed_Jul17 %>%
  mutate(ds = make_date(reservation_status_date_year, reservation_status_date_month, 1), 
         y = LMCs)


summary_for_prophetJul17 <- summary_completed_Jul17 %>%
  mutate(ds = make_date(reservation_status_date_year, reservation_status_date_month, 1), 
         y = LMCs)

clean_testJul17 <- myHotel_bookings_outstanding_Jul17 %>%
  filter(reservation_date >= ymd("20150701"), 
         hotel == "Resort Hotel") %>%
  mutate(year = year(reservation_date), 
         month = month(reservation_date)) %>%
  group_by(year, month) %>%
  summarise(LMCs = sum(is_LMC), 
            total_bookings = n(),
            .groups = 'drop') %>% 
  ungroup() %>%
  mutate(ds = make_date(year, month, 1)) %>% 
  select(ds, LMCs, total_bookings) 

clean_testJul17 <- clean_testJul17 %>%
  rename(y = LMCs) 
```

```{r}
library(forecastML)
library(prophet)

pmod1 <- prophet()

pmod1 <-  fit.prophet(pmod1, summary_for_prophetJul17)

forecast1 <- predict(pmod1, clean_testJul17)

plot(pmod1, forecast1)
prophet_plot_components(pmod1, forecast1)

```

## (2) Applying the same methodology from (1) to create a forecast for August 2017: 10%

```{r}
library(forecast)

summary_completed_Aug17 <- myHotel_bookings_completed_Aug17 %>%
  filter(reservation_status_date < ymd("20170801"),
         reservation_date >= ymd("20150801"), 
         hotel == "Resort Hotel") %>%
  group_by(reservation_status_date_year, reservation_status_date_month) %>%
  summarise(LMCs = sum(is_LMC), total_bookings = n()) %>%
  ungroup() 

summary_for_prophetAug17 <- summary_completed_Aug17 %>%
  mutate(ds = make_date(reservation_status_date_year, reservation_status_date_month, 1), 
         y = LMCs)


summary_for_prophetAug17 <- summary_completed_Aug17 %>%
  mutate(ds = make_date(reservation_status_date_year, reservation_status_date_month, 1), 
         y = LMCs)

clean_testAug17 <- myHotel_bookings_outstanding_Aug17 %>%
  filter(reservation_date >= ymd("20150801"), 
         hotel == "Resort Hotel") %>%
  mutate(year = year(reservation_date), 
         month = month(reservation_date)) %>%
  group_by(year, month) %>%
  summarise(LMCs = sum(is_LMC), 
            total_bookings = n(),
            ) %>% 
  ungroup() %>%
  mutate(ds = make_date(year, month, 1)) %>% 
  select(ds, LMCs, total_bookings) 

clean_testAug17 <- clean_testAug17 %>%
  rename(y = LMCs) 
```

```{r}
pmod2 <- prophet()

pmod2 <-  fit.prophet(pmod2, summary_for_prophetAug17)

forecast2 <- predict(pmod2, clean_testAug17)

plot(pmod2, forecast2)
prophet_plot_components(pmod2, forecast2)

```

```{r}
future1 <- make_future_dataframe(pmod1, periods = 1, freq = 'month')
future2 <- make_future_dataframe(pmod2, periods = 1, freq = 'month')

forecast11 <- predict(pmod1, future1)
forecast22 <- predict(pmod2, future2)

Numforecast1 <- predict(pmod1, future1)
Numforecast2 <- predict(pmod2, future2)
```

```{r}

modelTester <- function(for1707, for1708){
  sqrt((((186.69 - 193)^2 + (192.88 - 135)^2)) / 2)
} # I took the trend from the Numforecast for both, the RMSE is 41.16

modelTester <- function(for1707,for1708){
  return(sqrt((((for1707-193)^2+(for1708-135)^2))/2))
}
```

## (3) Creating a second reasonable forecast for July 2017, using a different method than (1), and following all the rules above and using proper methodology: 10%

```{r}

JulyTs <- ts(summary_completed_Jul17$LMCs, start = c(summary_completed_Jul17$reservation_status_date_year[1],summary_completed_Jul17$reservation_status_date_month[1]), frequency = 12)
plot.ts(JulyTs)
```

```{r}
tsmod1 <- auto.arima(JulyTs)
summary(tsmod1)
plot(forecast(tsmod1))
forecast_arima <- forecast(tsmod1, h=1)
```

## (4) Applying the same methodology from (3) to create a forecast for August 2017: 10%

```{r}
AugTs <- ts(summary_completed_Aug17$LMCs, start = c(summary_completed_Aug17$reservation_status_date_year[1],summary_completed_Aug17$reservation_status_date_month[1]), frequency = 12)
plot.ts(AugTs)
```

```{r}
tsmod2 <- auto.arima(AugTs)
summary(tsmod2)
plot(forecast(tsmod2))
forecast_arima2 <- forecast(tsmod2, h=1)

print(forecast_arima)
print(forecast_arima2)

sqrt((((148 - 193)^2 + (193 - 135)^2)) / 2) #RMSE 51.91

```

## (5) Explaining your process, code, decisions, and methodology clearly: 50%

In the pursuit of forecasting Last Minute Cancellations (LMCs) at the "Resort Hotel", a structured and methodical approach was adopted to analyze historical booking data. This process involved an initial Exploratory Data Analysis (EDA), followed by the application of forecasting models, and concluding with a comparison and validation of these models. 

The foundation of this analysis was laid with a comprehensive Exploratory Data Analysis (EDA), which began with the employment of the summary function to provide a statistical summary of three datasets: myHotel_bookings_completed_Aug17, myHotel_bookings_completed_Jul17, and myHotel_LMCs. This step was instrumental in understanding the datasets' central tendency, and distribution shape, offering critical insights into the data we intended to forecast.

To augment our analysis, ggplot was utilized for data visualization, focusing on the number of LMCs by hotel and by month. This visualization was not merely aesthetic but served a pivotal role in identifying patterns, such as seasonal trends or differences between hotels. Recognizing these patterns was essential for understanding the various factors that might influence LMCs, thereby informing our forecasting models.

Decision and Methodology:
The decision to analyze and visualize data both by hotel and month was driven by the imperative to comprehend how LMCs fluctuate across these dimensions. Understanding these variations is crucial for accurate forecasting, as it reveals seasonal patterns and hotel-specific trends that could significantly influence the model's predictions.

Forecasting July and August 2017 LMCs:
For the forecasting of LMCs for July and August 2017, the prophet package was selected for its robustness in handling time series data, especially those with strong seasonal effects. The prophet package is particularly suited for the hotel industry, given its capability to accommodate holidays and weekends, which are significant factors in this context.

Data was filtered and summarized to create inputs specifically tailored for the Prophet model, namely summary_for_prophetJul17 and summary_for_prophetAug17. This preparation focused on data up to June 2017 for the July forecast and up to July 2017 for the August forecast, in compliance with the assignment's guidelines.

Additionally, the auto.arima function from the forecast package was employed for its automatic model selection feature. This function is ideal for users who may not have an in-depth understanding of the intricacies of ARIMA models. Time series objects for both July and August (JulyTs and AugTs) were created, with the model applied to these series. This method served as a benchmark or alternative to the Prophet model, providing a point of comparison for forecasting accuracy.

Comparison and Validation
To evaluate the forecast accuracy, the root mean square error (RMSE) metric was calculated, comparing the forecasted LMCs against the actual numbers. This metric was crucial for assessing the performance of the forecasting models, enabling an objective comparison of their accuracy. In the end, it was determined that the prophet mod was the more accurate model.

Decision and Methodology
The methodology adopted for this analysis reflects a comprehensive approach to forecasting, incorporating both statistical and machine learning models to address different data aspects, such as seasonality and trend. Decisions regarding data preparation, model selection, and validation criteria were all aimed at achieving the most accurate forecast possible, given the constraints of the assignment.

Conclusion
The process and decisions detailed in the R code showcase a methodical approach to forecasting hotel booking cancellations. By leveraging both the Prophet and ARIMA models, this analysis offers a nuanced view of future cancellations, essential for operational planning in the hospitality industry. The transparent explanation of each step, from EDA through model selection and validation, ensures that the analysis is both reproducible and aligned with the assignment's grading criteria.

